A formal treatment of bidirectional typing

### 2023-10-11 tex/ESOP24/S1-intro.tex
[ update ] Add the missing .bib

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
\citet{Dunfield2021} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
While it is straightforward to derive a type synthesis algorithm,
For bidirectional typing, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
\LT{emphasise that formalisation and verification are important}

To remedy the situation, we develop a general and formal treatment of bidirectional typing, providing an out-of-the-box \emph{verified generator} of \emph{proof-relevant} type synthesisers developed with the proof assistant \Agda:

\subsubsection{Proof-relevance in (bidirectional) type synthesis}
\label{sec:PLFA}

@@

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in type theory, and we say $P$~is logically \emph{decidable} if the law holds for~$P$.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial.
Given that all proofs as programs terminate, logical decidability implies algorithmic decidability.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
It yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Then, both algorithmic soundness and completeness in original form are implied.
The proof of this statement in type theory is a \emph{proof-relevant} type synthesiser, because it does not only return yes/no but construct a proof---either a typing derivation or its refutation. 
\LT{emphasise that our proof is an actual program}
%For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.

@@

\subsubsection{Mode-correctness and a general definition of typed languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
It seems that the implications of mode-correctness have only been treated informally so far as exhibited in the survey paper~\cite{Dunfield2021}, since the notion of mode-correctness is not formally defined for \emph{languages} there.
%But, what is a language anyway?

To make the notion of mode-correctness precise, we exhibit a general definition of bidirectional simple type systems which extends the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
Such a general definition allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the logical decidability of bidirectional type synthesis for \emph{all} mode-correct \emph{bidirectional binding signatures}.
The latter theorem effectively amounts to a verified proof-relevant bidirectional-type-synthesiser generator for all syntax-directed simply typed languages.

To make our exposition accessible we have chosen to work with simply typed languages so that bidirectional type synthesis is established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
Nevertheless, we briefly discuss another definition which includes polymorphically typed languages such as \SystemF and additional assumptions we need in \cref{sec:future}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness,
  \item the concept of mode decoration and the comparison between our completeness and \varcitet{Dunfield2021}{'s annotatability}, 
    \Josh{generalisation mode decoration?}
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular.
\LT{methodology for combining proofs and programs?}

>>>>

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
\varcitet{Dunfield2021}{ in their survey paper} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
While it is straightforward to derive a type synthesis algorithm,
for bidirectional typing, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

Implementing a type synthesiser correctly, especially for advanced type systems, is non-trivial.
For example, proof assistants in use including \Coq and \Agda have many bugs in their type synthesis such as absurdity, supposedly equal terms which were not normalised to the same canonical term, and so on.
\LT{some GitHub issues \cite{Felicissimo2023}}
The gap between the design of type synthesis and its implementation cannot be ignored.

To mitigate the above problem, we demonstrate a general and formal treatment of bidirectional typing, providing an out-of-the-box \emph{verified generator} of \emph{proof-relevant} type synthesisers developed with the proof assistant \Agda:

\subsubsection{Proof-relevance in type synthesis}
\label{sec:PLFA}

@@

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in type theory, and we say $P$~is logically \emph{decidable} if the law holds for~$P$.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial.
In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
It yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Hence, both algorithmic soundness and completeness in original form are implied.
On the other hand, the proof of this statement in type theory is a \emph{proof-relevant} type synthesiser, because it does not only return yes/no but construct a proof---either a typing derivation or its refutation. 
\LT{emphasise that our proof is an actual program}
%For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.

@@

\subsubsection{Mode-correctness and general definitions of languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
It seems that the implications of mode-correctness have only been addressed informally as exhibited in the survey paper~\cite{Dunfield2021}, since mode-correctness is not formally defined as a \emph{property of languages}.
%But, what is a language anyway?

To make the notion of mode-correctness precise, we give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, which extends the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
Such a general definition allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct bidirectional binding signatures.
The latter theorem effectively amounts to a verified proof-relevant bidirectional-type-synthesiser generator for all syntax-directed bidirectional simple type systems that are mode-correct.

To make our exposition accessible we have chosen to work with simply typed languages so that bidirectional type synthesis is established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
Nevertheless, we briefly discuss a possible extension which includes polymorphically typed languages such as \SystemF with additional assumptions in \cref{sec:future}.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness,
  \item the concept of mode decoration and the comparison between our completeness and \varcitet{Dunfield2021}{'s annotatability}, 
    \Josh{generalisation mode decoration?}
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular, filing the gap between theory and implementation.
